---
title: "Assignment data analysis in practice 202307243"
author: "Rachel Andine"
date: "2025-04-21"
output: html_document
---
#import csv file to R and prepare data 

In this step, we will load the wine quality dataset, explore its basic structure, and prepare it for further analysis. This includes:
- Importing the data
- Checking its structure and summary
- Identifying any missing values or duplicates
- Scaling (standardizing) the numeric variables for modeling

To help us in our task of cleaning the data wer are using the tidyverse library 

The dataset contains physicochemical attributes of red wine samples along with a sensory quality rating (from 0 to 10). The table below summarizes the distribution of each variable using key descriptive statistics: minimum, 1st quartile, median, mean, 3rd quartile, and maximum.

🔍 Key Observations:
Fixed Acidity: Ranges from 4.6 to 15.9, with a mean of 8.32. Most values fall between 7.1 and 9.2.

Volatile Acidity: Skews low, ranging from 0.12 to 1.58. The median is 0.52, showing most samples are on the lower side, which is desirable for wine quality.

Citric Acid: Mostly small values (median = 0.26), with a few outliers up to 1.0.

Residual Sugar: The range (0.9 to 15.5) shows high variability; the mean (2.54) is higher than the median (2.2), suggesting a right-skewed distribution due to outliers.

Chlorides: Varies between 0.012 and 0.611, with a median of 0.079, and a slight positive skew.

Free Sulfur Dioxide: Median of 14, but a max of 72, indicating some extreme values.

Total Sulfur Dioxide: Widely dispersed, ranging up to 289 with a median of 38. This could impact flavor and preservation.

Density: Very narrow range (0.9901 to 1.0037), as expected for liquids, with a mean of 0.9967.

pH: Ranges from 2.74 to 4.01, centered around a mean of 3.31 — typical acidity for red wine.

Sulphates: Ranges from 0.33 to 2.00; most values lie below 0.73, with a positively skewed distribution.

Alcohol: Varies from 8.4% to 14.9%. The mean (10.42%) suggests moderate alcohol content overall, with higher alcohol levels possibly improving wine quality.

Quality (Target Variable): Scores range from 3 to 8, with a median and mode of 6. The average score is 5.64, indicating a generally moderate quality level, with some outliers of both low and high quality.

while checking the data set for missing values turn out we had no missing values as they returned 0 for all column 

We have found 240 duplicate acrross all rows which we have later delete to keep only the unique rows 



Using the IQR method, we identified a notable number of outliers across several numeric variables. These outliers are values that fall significantly below or above the typical range of the data.

Notably:

residual_sugar (126 outliers) and chlorides (87 outliers) show substantial deviation, which may affect modeling if left untreated.

citric_acid had only 1 outlier, and alcohol had 12, indicating more stable distributions.

The presence of outliers in variables such as sulphates, fixed_acidity, and volatile_acidity should be assessed visually to determine whether they represent data quality issues or genuine extreme values.

```{r}
# Get tidyverse package Essential for data manipulation and visualization, includes packages like ggplot2, dplyr (data manipulation tasks such as filtering rows, selecting columns, and arranging data), and tidyr.
library(tidyverse)
library(janitor)

# Read the CSV file into R and store it in a variable called 'wine'

wine <- read_delim("Wine Quality (Red Wine).csv", delim = ";")
#view(wine) #to view the whole data set 

#display the first rows of the wine dataset and the variable types 
head(wine)
skimr::skim(wine)
glimpse(wine)  # shows all column names and types
summary(wine)

# Clean column names
wine <- wine %>% clean_names()

# Check for missing values
missing_total <- sum(is.na(wine))
if (missing_total == 0) {
  print("✅ No missing values in the dataset.")
} else {
  print(paste("⚠️", missing_total, "missing values found."))
}

# Check for duplicate rows
duplicate_total <- sum(duplicated(wine))
if (duplicate_total == 0) {
  print("✅ No duplicated rows found.")
} else {
  print(paste("⚠️", duplicate_total, "duplicated rows detected."))
}
# This removes duplicated rows across all columns
wine <- wine %>% distinct()

#to check if duplicate was removed 
skimr::skim(wine)
```


``` {r}

# Reshape the data to long format for easy faceting
wine_long <- wine %>%
  pivot_longer(cols = -quality, names_to = "variable", values_to = "value")

# Plot histograms for all variables (except quality)
ggplot(wine_long, aes(x = value)) +
  geom_histogram(fill = "#2E86C1", bins = 30, color = "white") +
  facet_wrap(~variable, scales = "free", ncol = 3) +
  labs(title = "Distribution of Numeric Features in Wine Dataset", x = "Value", y = "Count") +
  theme_minimal()

#to check for outlier 
detect_outliers <- function(x) {
  iqr <- IQR(x)
  lower <- quantile(x, 0.25) - 1.5 * iqr
  upper <- quantile(x, 0.75) + 1.5 * iqr
  sum(x < lower | x > upper)
}

sapply(wine %>% select(-quality), detect_outliers)
# Load ggplot2 (already included in tidyverse)
library(ggplot2)

# Create a long-format version of the data
wine_long <- wine %>%
  pivot_longer(cols = -quality, names_to = "variable", values_to = "value")

# Boxplots for selected variables with high outliers
high_outlier_vars <- c("residual_sugar", "chlorides", "sulphates", "fixed_acidity")

ggplot(wine_long %>% filter(variable %in% high_outlier_vars), 
       aes(x = variable, y = value)) +
  geom_boxplot(fill = "#2E86C1", outlier.color = "red") +
  labs(title = "Boxplots for High Outlier Features", y = "Value", x = "Feature") +
  theme_minimal()

